Aligning marine species range data to better serve science and conservation

### 2016-03-10 manuscript-main.Rmd
updating manuscript

---
title: "Aligning marine species range data to better serve science and conservation"
author: "Casey O'Hara^1^, Jamie Afflerbach^1^, Benjamin S. Halpern<sup>1, 2, 3</sup>"
output:
  html_document:
    highlight: haddock
    number_sections: yes
    theme: spacelab
    toc: no
  pdf_document:
    number_sections: yes
    toc: no
  word_document: default
---

>>>>

---
title: "Aligning marine species range data to better serve science and conservation"
author: "Casey O'Hara^1^, Jamie Afflerbach^1^, Benjamin S. Halpern<sup>1, 2, 3</sup>, Courtney Scarborough^1^"
output:
  html_document:
    highlight: haddock
    number_sections: yes
    theme: spacelab
    toc: no
  pdf_document:
    number_sections: yes
    toc: no
  word_document: default
---

### 2016-03-10 mpa_analysis/3_mpa_spp_analysis.Rmd
created IUCN spp cell summary to identify gap spp and 10%+ spp
updating manuscript

### compare results using only IUCN species maps
dir_iucn_data <- '/var/data/ohi/git-annex/globalprep/_raw_data/iucn_spp'
iucn_cells_file <- file.path(dir_iucn_data, 'iucn_cells_v2016.csv')
iucn_spp_cells <- read_csv(iucn_cells_file, col_types = 'cddddc') %>%
  as.data.frame(stringsAsFactors = FALSE)

>>>>

### compare results using only IUCN species maps
dir_iucn_data <- file.path(dir_N, 'git-annex/globalprep/_raw_data/iucn_spp')
iucn_cells_file <- file.path(dir_iucn_data, 'iucn_cells_v2016.csv')
iucn_spp_cells <- read_csv(iucn_cells_file, col_types = 'cddddc') %>%
  as.data.frame(stringsAsFactors = FALSE)

### 2016-03-10 ms_2_intro_res_disc.Rmd
updating manuscript

The methodologies behind the creation of these datasets imply differences in prediction of species distribution due to errors of commission (falsely indicating species presence) and omission (falsely indicating species absence).  Geographic range data such as IUCN range maps frequently introduce commission errors, while species distribution models such as AquaMaps will likely introduce fewer commission errors but more omission errors.  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range (Rondinini et al. 2006).  ___By comparing maps resulting from each methodology, we can identify and possibly address mechanistic causes for each type of error.___

<!--BSH: I haven’t read the paper, but it sounds like we need to acknowledge what was already discussed/done there.  This may actually simplify things, such that we condense the first three paragraphs of this section into a single paragraph that restates what that paper found (and cite it) and add anything else we want to point out.  But we will definitely need a succinct statement about how our paper moves well beyond what that paper did. Not sure best approach without reading the paper.-->

@@

To explore differences in species distribution and range between the two datasets, we plotted the distribution alignment (how much of the smaller range falls within the larger range, i.e. where on the map) against the area alignment (ratio of smaller range area to larger range area, i.e. how much of the map) for each shared species (Fig. 3A). 
 This analysis revealed a general negative linear pattern, suggesting that increasing similarity in range area correlates with decreasing distribution alignment.
 AquaMaps tends to extrapolate species ranges beyond known occurrences, such that each additional unit of range predicted by AquaMaps will fall in different locations than an additional unit of range predicted using IUCN methodology. For species with dissimilar range areas, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar range areas, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

The mean distribution alignment for species included in both datasets was `r dist_align_mean`%; the mean area alignment was `r area_align_mean`%. By dividing the map-paired species into quadrants based on these means, we highlight categories of relationships that help further explain this general pattern. The upper right quadrant includes species (n = `r n_q1`) whose described ranges are above average in both spatial distribution and extent. These species tend to be well-studied and include wide-ranging pelagic organisms such as marine mammals, tunas, and billfishes (Fig. 3B). This result is not surprising, as species with very large ranges are likely to be more aligned regardless of methodology simply because their ranges span nearly the entire map.

The area-mismatched ranges contained in the upper left quadrant (n = `r n_q2`) include many species whose spatial distribution is similar, but where the IUCN range is notably larger, often extending into deeper water.  For example, corals dominate this quadrant (n = `r n_q2_corals`; `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in this quadrant), and IUCN range maps tend to extend corals into waters beyond their preferred depths, likely introducing errors of commission. ___See supplemental information for maps that highlight this issue.___ _should there just be a blanket "see SOM for maps that typify each quadrant"?_
Ocean depth is explicitly included in AquaMaps models, while depth is frequently overlooked as a factor in IUCN range maps.  Simply clipping IUCN range maps to known depth preferences would resolve many of these mismatches.

@@

___While we examined IUCN Red List extinction risk categories by quadrant, we found no compelling patterns.  See SOM for details.___

@@

Clear and significant differences in species range descriptions between AquaMaps and IUCN range maps due to method-driven differences in commission and omission errors.  To examine the implications of these differences, we replicated two recent studies, varying only the prioritization of one data set over the other.

@@

___Should end with a more solid statement of how choosing one or the other datasets/thresholds would have different results – using our calculated differences.  JA___

@@

@@

__Suggestions for data providers:__

* The IUCN Red List instructional materials for mapping marine species suggest that "bathymetry can be used to delineate species' ranges limited by depth in the same way as elevation is used for terrestrial species" (IUCN 2015).  Unfortunately, this recommendation appears to be applied inconsistently.  Especially for demersal communities dependent on photosynthesis, bathymetry should be a primary consideration; even a simple cutoff at the photosynthetic limit of 200 meters would minimize a likely source of commission errors.
* The AquaMaps model is a powerful tool for quickly estimating species distribution based on limited information; however, the model's output is only as good as the input data, which for marine species is often sorely limited.  AquaMaps encourages experts to review and comment on the predicted species distribution maps; lessons from the small sample of reviewed maps could be generalized, for example to other species in the same taxonomic group, to further refine the predictions.  This is especially important for data-poor species.
* In addition to environmental preferences and conditions, AquaMaps mapping parameters include area restrictions to help limit over-extrapolation of AquaMaps models.  Currently area restrictions are denoted by UN Food and Agriculture Organisation (FAO) fisheries reporting areas (Ready et al. 2010); however, area restrictions based on biogeographical criteria, e.g. Marine Ecoregions of the World (Spalding et al. 2007), would likely provide better resolution and predictive power, especially for data-poor species.

__Suggestions for data users:__

* For depth-limited species and taxa, clipping IUCN range polygons to a reasonable bathymetry contour can reduce commission errors due to overprediction of species presence into unsuitable habitats.
* Using both datasets together can increase the taxonomic and spatial breadth of coverage, as long as the differences between the datasets can be reasonably minimized.  Consider:
    * An AquaMaps presence threshold of 0% most closely approximates the "limits of distribution" defined by IUCN range maps.
    * Additional filters on AquaMaps distribution criteria, such as expert review status or the number of occurrence cells used to generate the species model, can help avoid overextrapolation for data poor species.

>>>>

The methodologies behind the creation of these datasets imply differences in prediction of species distribution due to errors of commission (falsely indicating species presence) and omission (falsely indicating species absence).  Geographic range data such as IUCN range maps frequently introduce commission errors, while species distribution models such as AquaMaps will likely introduce fewer commission errors but more omission errors.  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range (Rondinini et al. 2006).  By comparing maps resulting from IUCN and AquaMaps methodolog, we can identify and possibly address mechanistic causes for each type of error.

@@

To explore differences in species distribution and range between the two datasets, we plotted the distribution alignment (how much of the smaller range falls within the larger range, i.e. where on the map) against the area alignment (ratio of smaller range area to larger range area, i.e. how much of the map) for each shared species (Fig. 3A). 
 This analysis revealed a general negative linear pattern, suggesting that increasing similarity in range area correlates with decreasing distribution alignment.
 AquaMaps tends to extrapolate species ranges into suitable areas beyond known occurrences, such that each additional unit of range predicted by AquaMaps will fall in different locations than an additional unit of range predicted using IUCN methodology. For species with dissimilar range areas, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar range areas, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

The mean distribution alignment for species included in both datasets was `r dist_align_mean`%; the mean area alignment was `r area_align_mean`%. By dividing the map-paired species into quadrants based on these means, we highlight categories of relationships that help further explain this general pattern. For representative maps from each category, please refer to the supplemental materials (SI Fig. XXX).

The upper right quadrant includes species (n = `r n_q1`) whose described ranges are above average in both spatial distribution and extent. These species tend to be well-studied and include wide-ranging pelagic organisms such as marine mammals, tunas, and billfishes (Fig. 3B). This result is not surprising, as species with very large ranges are likely to be more aligned regardless of methodology simply because their ranges span nearly the entire map.

The area-mismatched ranges contained in the upper left quadrant (n = `r n_q2`) include many species whose spatial distribution is similar, but where the IUCN range is notably larger, often extending into deeper water.  For example, corals dominate this quadrant (n = `r n_q2_corals`; `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in this quadrant), and IUCN range maps tend to extend corals into waters beyond their preferred depths, likely introducing errors of commission  (SI Figs. XXX, XXX). 
 Ocean depth is explicitly included in AquaMaps models, while depth is frequently overlooked as a factor in IUCN range maps.  Simply clipping IUCN range maps to known depth preferences would resolve many of these mismatches.

@@

While we examined IUCN Red List extinction risk categories by quadrant, expecting to see more concurrence between maps for species of concern; however, we found no compelling patterns (SI Fig. XXX).

@@

Method-driven differences in commission and omission errors drive clear and significant differences in species range descriptions between AquaMaps and IUCN datasets.   To examine the implications of these differences, we replicated two recent studies, varying only the prioritization of one data set over the other.

@@

@@

- check which species are gap species and look at some maps
- check which species are 10%+ species and look at some maps
- where do these species fall georegionally? is there anything odd about their ranges?

@@

The IUCN Red List instructional materials for mapping marine species suggest that "bathymetry can be used to delineate species' ranges limited by depth in the same way as elevation is used for terrestrial species" (IUCN 2015).  Especially for demersal communities dependent on photosynthesis, bathymetry should be a primary consideration; data users might consider a simple cutoff at the photosynthetic limit of 200 meters to minimize a likely source of commission errors.

The AquaMaps methodology is a powerful means of quickly estimating species distribution based on limited information; however, the model's output is only as good as the input data, which for marine species is often sorely limited.  In addition to environmental preferences and conditions, AquaMaps mapping parameters include area restrictions to help limit over-extrapolation of AquaMaps models.  Currently, these area restrictions are denoted by UN Food and Agriculture Organisation (FAO) fisheries reporting areas (Ready et al. 2010); however, area restrictions based on biogeographical criteria, e.g. Marine Ecoregions of the World (Spalding et al. 2007), would likely decrease commission errors and improve predictive power, especially for data-poor species.

When using AquaMaps data alone, changing the probability of occurrence threshold to determine presence becomes a tradeoff between errors of commission (low threshold) and omission (high threshold).  Using AquaMaps in conjunction with IUCN can increase the taxonomic and spatial breadth of coverage, as long as the differences between the datasets can be reasonably minimized. In this case, we recommend a presence threshold of 0% as it most closely approximates the "limits of distribution" criterion defined by IUCN data providers.

<!-- * Additional filters on AquaMaps distribution criteria, such as expert review status or the number of occurrence cells used to generate the species model, can help avoid overextrapolation for data poor species. _using other attributes and variables to filter? this seems outside the scope of this paper_ -->

### 2016-03-10 ms_3_methods.Rmd
updating manuscript

**Comparison of taxonomic and regional distribution:** 
To examine the overall taxonomic distribution across the spatial datasets (Fig. 1A), we grouped species by taxonomic class and data source (IUCN, AquaMaps, or both), and examined the proportion of each class represented in each data source category.

To compare the spatial representation of the two datasets directly, we rasterized the IUCN species polygons to the same 0.5° grid as the AquaMaps species maps; species presence within a grid cell was determined by any non-zero overlap of a species polygon with the cell, and species richness per cell was simply the count of the species present.  For the AquaMaps dataset, we determined per-cell species richness by counting all species with non-zero probability of occurrence, to best approximate the "extent of occurrence" generally indicated by IUCN maps.  We represented relative distribution of species richness for each dataset by plotting average species count against latitude and longitude (Fig. 1B, 1C).

@@

Overlaying paired distribution maps for a given species, we defined and calculated _distribution alignment_ and _extent alignment_ and plotted these in Fig. 2:

@@

We visually inspected a random selection of paired distribution maps from each quadrant to identify possible mechanistic causes of misalignment.  To verify that IUCN predicted unsuitable habitat for depth limited species, we used QGIS (REF) to overlay a selection of IUCN and AquaMaps maps with a 200 meter bathymetry contour.

>>>>

**Comparison of taxonomic and regional distribution:** 
To examine the overall taxonomic distribution across the spatial datasets (Fig. 1), we grouped species by taxonomic class and data source (IUCN, AquaMaps, or both), and examined the proportion of each class represented in each data source category.

To compare the spatial representation of the two datasets directly, we rasterized the IUCN species polygons to the same 0.5° grid as the AquaMaps species maps; species presence within a grid cell was determined by any non-zero overlap of a species polygon with the cell, and number of species per cell was simply the count of the species present.  For the AquaMaps dataset, we determined per-cell species count by including all species with non-zero probability of occurrence, to best approximate the "extent of occurrence" generally indicated by IUCN maps.  We represented relative distribution of species count for each dataset by plotting average species count against latitude and longitude (Fig. A, 2B).

@@

Overlaying paired distribution maps for a given species, we defined and calculated _distribution alignment_ and _extent alignment_ and plotted these in Fig. 3A:

@@

For each species with ranges described in both IUCN and AquaMaps, $A_{small}$ and $A_{large}$ indicate the smaller and larger range representation (regardless of which dataset).  $A_{small \cap large}$ represents the amount of overlapping area between the two datasets. We visually inspected a random selection of paired distribution maps from each quadrant to identify possible mechanistic causes of misalignment.  To verify that IUCN predicted unsuitable habitat for depth limited species, we used QGIS (REF) to overlay a selection of IUCN and AquaMaps maps with a 200 meter bathymetry contour.
